---
title: "1. Volcano plot"
author: "Runshan Gui"
date: "`r Sys.Date()`"
output: html_document
---

### 1.Clear the Environment
```{r}
rm(list=ls())
```

### 2.Loading Required Packages for Volcano Plot
```{r}
library(readxl)
library(DESeq2)
library(ggplot2)
library(tibble)
library(ggrepel)
library(tidyverse)
library(dplyr)
library(patchwork)
```

### 3.Reading the Excel File
```{r}
gene_expression_annotation <- read_xlsx("gene_expression_annotation.xlsx")
```

### 4.Extracting Counts Data and Creating an Integer Matrix
```{r}
count_columns <- c("CON1_count",
                   "CON2_count",
                   "CON3_count",
                   "ICI-AM_1_count",
                   "ICI-AM_2_count",
                   "ICI-AM_3_count"
                   )
count_data <- gene_expression_annotation[, count_columns]
count_matrix <- as.matrix(count_data)
mode(count_matrix) <- "integer"    ## Ensure the data is of integer type
rownames(count_matrix) <- gene_expression_annotation$gene_id    ## Set gene IDs as row names
```

### 5.Creating a Sample Information Data Frame
```{r}
sample_info <- data.frame(
  condition = rep(c("Control", "ICI_AM"), each = 3),
  row.names = colnames(count_matrix)
  )
```

### 6.Creating a DESeqDataSet Object
```{r}
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = sample_info,
  design = ~ condition
  )
```

### 7.Filtering Low-Expression Genes and Running DESeq2 Differential Analysis
```{r}
## Counts ≥ 1 in at least 2 samples
keep <- rowSums(counts(dds) >= 1) >= 2
dds <- dds[keep, ]
dds <- DESeq(dds)
```

### 8.Extracting Analysis Results (ICI_AM vs Control)
```{r}
results <- results(dds,
                   contrast = c("condition",
                                "ICI_AM",
                                "Control"
                                )
                   )
```

### 9.Converting Results to a Data Frame and Adding Gene Annotations
```{r}
res_df <- as.data.frame(results)
res_df$gene_id <- rownames(res_df)
final_results <- merge(res_df,
                       gene_expression_annotation[, c("gene_id",
                                                      "gene_name",
                                                      "description"
                                                      )],
                       by = "gene_id",
                       all.x = TRUE
                       )
```

### 10.Sorting by p-value and Viewing Significant Results
```{r}
final_results <- final_results[order(final_results$pvalue), ]
significant_genes <- subset(final_results, padj < 0.05)
head(significant_genes)
```

### 11.Saving Results to a CSV File件
```{r}
write.csv(final_results, "res_df.csv", row.names = FALSE)
```

### 12.Plotting the Volcano Plot
```{r}
## 1.Define upregulated and downregulated differentially expressed genes
final_results$diffexpressed <- "Not Significant"
final_results$diffexpressed[final_results$log2FoldChange > 0.5 & final_results$padj < 0.15] <- "up"
final_results$diffexpressed[final_results$log2FoldChange < -0.5 & final_results$padj < 0.15] <- "down"

## 2.Calculate -log10(padj) for subsequent visualization reference
final_results$logP <- -log10(final_results$padj)

## 3.Extract significant or non-significant points
data_bg <- final_results[final_results$diffexpressed == "Not Significant", ]
data_sig <- final_results[final_results$diffexpressed != "Not Significant", ]

## 4.Add labels for top genes
# Significantly differentially expressed genes
highlight_genes <- c("Trim47", "Neurl3")
data_label <- final_results[final_results$gene_name %in% highlight_genes, ]

## 5.Color settings
colors <- c("up" = "#fe8264",
            "down" = "#b5dbe6",
            "normal" = "#474747",
            "normalIn" = "#474747"
            )

## 6.Plotting with ggplot2
p <- ggplot(final_results,
            aes(x = log2FoldChange,
                y = logP)) +
  geom_point(data = data_bg,
             shape = 21,
             fill = "#474747",
             color = "black",
             alpha = 0.8,
             size = 1.3,
             stroke = 0.3) +
  geom_point(data = data_sig,
             aes(fill = diffexpressed),
             shape = 21,
             color = "black",
             size = 2,
             stroke = 0.3) +
  scale_fill_manual(name = "Expression",
                    values = colors,
                    ) +
  geom_text_repel(data = data_label,
                  aes(x = log2FoldChange,
                      y = logP,
                      label = gene_name),
                  size = 12,
                  force_pull = 0,
                  point.padding = 0,
                  box.padding = 1,
                  min.segment.length = 0,
                  vjust = 0.5,
                  segment.color = "grey20",
                  segment.size = 0.5,
                  segment.alpha = 0.8,
                  max.overlaps = Inf,
                  seed = 42,
                  max.time = 1,
                  max.iter = Inf,
                  nudge_y = 0.5,
                  nudge_x = -0.3) +
  geom_hline(yintercept = -log10(0.15),
             color = "#474747",
             linewidth = 0.6,
             linetype = "dashed") +
  scale_x_continuous(limits = c(-3, 3),
                     breaks = seq(-3, 3, 1)) +
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 6),
                     breaks = seq(0, 6, 1)) +
  labs(title = "Bulk RNA-seq (Con vs ICI-AM)",
       x = expression(log[2](FoldChange)),
       y = expression(-log[10](adj.p))) +
  theme_classic() +
  theme(legend.position = "right",
        legend.title = element_text(size = 30),
        legend.text = element_text(size = 30),
        plot.title = element_text(size = 36,
                                  face = "bold",
                                  hjust = 0.5),
        axis.title.x = element_text(size = 36),
        axis.title.y = element_text(size = 36),
        axis.text.x = element_text(size = 24),
        axis.text.y = element_text(size = 24)
        )
## 7.Save the plot
ggsave(filename = "Fig.2b.pdf", width = 12, height = 12, plot = p, bg = "white")
```
